Precision Fundamental Stellar Properties with Adaptive Machine Learning

### 2024-03-06 paper/paper2/ms.tex
Complete interpolation code and test dropout residual

\begin{document}
\title{Precision Fundamental Stellar Properties with Interpretable Machine Learning}

@@

Precomputed synthetic stellar spectral models represent a distillation of all available physics
and chemistry theory into effortlessly-reproducible comparisons for spectral observations.
Practitioners may simply download a file and overplot it against an observed spectrum, relying on
human pattern recognition to obtain insights on fundamental stellar properties. But such exercises
typically fall short. Inevitably, individual line depth and shape predictions do not match the
observed spectrum. \citet{2022ApJ...941..200G} introduced a two-stage interpretable machine learning (ML)
technique to alleviate some of these challenges. First, cloning the lines in a precomputed spectrum
with fine-tuned Voigt profiles, and then warping their line properties in response to data using
the power of automatic differentiation. This design satisfied a common use case, the desire to
obtain a semi-empirical super-resolution stellar template for the observed star: the pristine
$R=\frac{\lambda}{\delta \lambda}\sim1,000,000$ emergent spectrum coming from a patch of the star,
having accounted for rotational broadening, instrumental resolution, and telluric artifacts.
We called this algorithm \emph{blas\'e}, available with the Python package \texttt{blase}.

This capability alone was useful for Extreme Precision Radial Velocity (EPRV), but a key limitation
of \emph{blas\'e} was its requirement to pre-select a grid point as the initial starting point. The code
introduced in this study bypasses this requirement by comparing spectral reconstructions to an input
spectrum in order to automatically ascertain the fundamental stellar parameters.

@@

For the purposes of this study, we use a slice of the PHOENIX grid where the alpha abundance
is set to zero. Within this slice, we allow the full range of $\Teff, \logg, \Z$, leaving us with 1,959
grid points, each with its own PHOENIX spectrum. The specific domain and
sampling of the grid points are shown in Figure $XX$. Our approach could feasibly be extended to
include alpha abundances, but given that the \texttt{gollum} interface does not yet support the access
of spectra with nonzero alpha abundances, as well as bearing a significant increase in computational
impact, we leave this for future work.

The \texttt{blase} code takes in any spectrum as input, and outputs a cloned
version of that spectrum. However, there is a hidden product to this
procedure, the state dictionary of the model itself, which can be written to a file. In this lies metadata about
where spectral lines were identified by \texttt{blase}, as well as those lines'
specific properties, namely amplitude, Gaussian shape, and Lorentzian shape.
In addition, \texttt{blase} also allows the line centers to shift slightly during
cloning, meaning that in addition to the initial line centers, we
also have access to the shifted line centers. This hidden metadata is what
formed the backbone for this study and allowed us to retrieve line-by-line data
that could then be interpolated.

The ability to store these models on disk in itself is quite novel, as it offers a reduction
in storage space while also being interpretable. What we effectively do is store the
generator for the data rather than the data itself, which is a new paradigm in the 
storage of model grids such as PHOENIX.
Using this method, we input the PHOENIX grid spectra into \texttt{blase},
which wrote the models to disk, and then accessed the state dictionaries of each clone. Once this data was retrieved from the state dictionaries
across the grid, it was then aggregated into a single \texttt{pandas} DataFrame, enabling
querying of the data by any property, be it a fundamental stellar parameter or the output
line properties.

>>>>

\begin{document}
\title{Precision Fundamental Stellar Properties with Adaptive Machine Learning}

@@

Precomputed synthetic stellar spectral models represent a distillation of all available physics
and chemistry theory into effortlessly-reproducible comparisons for spectral observations.
Practitioners may simply download a file and overplot it against an observed spectrum, relying on
human pattern recognition to obtain insights on fundamental stellar properties. But such exercises
typically fall short. Inevitably, individual line depth and shape predictions do not match the
observed spectrum. \citet{2022ApJ...941..200G} introduced a two-stage interpretable machine learning (ML)
technique to alleviate some of these challenges: first, cloning the lines in a precomputed spectrum
with fine-tuned Voigt profiles, and then warping their line properties in response to data using
the power of automatic differentiation. This design satisfied a common use case, the desire to
obtain a semi-empirical super-resolution stellar template for the observed star: the pristine
$R\equiv\frac{\lambda}{\delta \lambda}\sim1,000,000$ emergent spectrum coming from a patch of the star,
having accounted for rotational broadening, instrumental resolution, and telluric artifacts.
We called this algorithm \emph{blas\'e}, available with the Python package \texttt{blase}.

This capability alone was useful for Extreme Precision Radial Velocity (EPRV), but a key limitation
of \emph{blas\'e} was its requirement to pre-select a single grid point as the initial starting point. The approach
introduced in this study bypasses this requirement by analyzing line properties of the input spectrum
in order to automatically ascertain fundamental stellar parameters.

@@

For the purposes of this study, we use a slice of the PHOENIX grid (Husser et al) where the alpha abundance
is set to zero. Within this slice, we choose the full range of $\Teff, \logg, \Z$, leaving us with 1,959
grid points, each with its own PHOENIX spectrum. The specific domain and
sampling of the grid points are shown in Figure $XX$.

The \texttt{blase} code takes in any spectrum as input, and outputs a cloned
version of that spectrum. There is a hidden product to this
procedure, the state dictionary of the model itself, which can be written to a file. In this lies metadata about
where spectral lines were identified by \texttt{blase}, as well as those lines'
specific properties, namely amplitude, Gaussian shape, and Lorentzian shape. This metadata is 
all of the information needed to reconstruct the spectrum.
In addition, \texttt{blase} also allows the line centers to shift slightly during
cloning, meaning that in addition to the initial line centers, we
also have access to the shifted line centers. This hidden metadata is what
formed the backbone for this study and allowed us to retrieve line-by-line data
that could then be interpolated.

The ability to store these models on disk in itself is quite novel, as it offers a reduction
in storage space while also being interpretable. What we effectively do is store the
generator for the data rather than the data itself, which is a new paradigm in the 
storage of model grids such as PHOENIX.
Using this method, we input 1,959 PHOENIX grid spectra into \texttt{blase},
which wrote the models to disk, and then accessed the state dictionaries of each clone. Once this data was retrieved from the state dictionaries
across the grid, it was then aggregated into a single table of shape $XX$, enabling
querying of the data by any property, be it a fundamental stellar parameter or the output
line properties.
